---
title: "Population estimate"
author: "Arfa Aijazi"
date: "November 2023"
output: html_document
---
This script estimates the population prevalence of temperature-related morbidity

### Setup
```{r}
library(survey)
library(tidyverse)
library(scales)
library(usmap)
library(ggpubr)
```

Custom plot theme
```{r message=FALSE, warning=FALSE, include=FALSE}
theme_custom = function() {
  theme_minimal() %+replace%
    theme(legend.position = "top") +
    theme(panel.grid = element_blank()) +
    theme(strip.background = element_blank()) +
    theme(panel.border = element_blank()) +
    theme(panel.background = element_blank()) +
    theme(legend.title = element_blank()) +
    theme(axis.title.y = element_blank()) +
    theme(text = element_text(size = 7, colour = "#000000")) +
    theme(plot.margin = unit(c(0,0,0,0), 'lines'))
}
```

Color palette
```{r}
## custom colors
my_pal <- rcartocolor::carto_pal(n = 4, name = "Pastel")
my_pal2 <- c(my_pal[1], my_pal[3])
```

### Load data
Load raw microdata files from the Energy Information Administration (EIA) website
```{r echo=TRUE, message=FALSE, warning=FALSE}
# Pull data from EIA website
# data_2015 = "https://www.eia.gov/consumption/residential/data/2015/csv/recs2015_public_v4.csv"
# data_2020 = "https://www.eia.gov/consumption/residential/data/2020/csv/recs2020_public_v5.csv"

# Predownloaded data file
data_2015 = "../data/recs2015_public_v4.csv"
data_2020 = "../data/recs2020_public_v5.csv"

recs_2015 <- read_csv(data_2015)
recs_2020 <- read_csv(data_2020)
```

2015 Survey analysis based on: https://www.eia.gov/consumption/residential/data/2015/pdf/microdata_v3.pdf
```{r}
sampweights_2015 <- recs_2015$NWEIGHT
brrwts <- recs_2015[,grepl("^BRRWT", names(recs_2015))]
des <- svrepdesign(weights = sampweights_2015, repweights = brrwts, type = "Fay", rho = 0.5, mse = TRUE, data = recs_2015)

HOTMA_2015 <- as.data.frame(svytotal(~HOTMA, des))
HOTMA_2015$CI <- HOTMA_2015$SE*1.96

COLDMA_2015 <- as.data.frame(svytotal(~COLDMA, des))
COLDMA_2015$CI <- COLDMA_2015$SE*1.96

results_2015 <- rbind(HOTMA_2015, COLDMA_2015) %>%
  rownames_to_column(var = "variable") %>%
  mutate(year = 2015)

```

2020 Survey analysis based on: https://www.eia.gov/consumption/residential/data/2020/pdf/microdata-guide.pdf
```{r}
repweights_2020 <- select(recs_2020, NWEIGHT1:NWEIGHT60)
recs_2020$POP_STATE <- 1
RECS <- svrepdesign(data = recs_2020,
                    weight = ~NWEIGHT,
                    repweights = repweights_2020,
                    type = "JK1",
                    combined.weights = TRUE,
                    scale = (ncol(repweights_2020)-1)/ncol(repweights_2020),
                    mse = TRUE)

HOTMA_2020 <- as.data.frame(svytotal(~HOTMA, RECS))
HOTMA_2020$CI <- HOTMA_2020$SE*1.96

COLDMA_2020 <- as.data.frame(svytotal(~COLDMA, RECS))
COLDMA_2020$CI <- COLDMA_2020$SE*1.96

results_2020 <- rbind(HOTMA_2020, COLDMA_2020) %>%
  rownames_to_column(var = "variable") %>%
  mutate(year = 2020)
```

2020 state level estimates
```{r}
POP_STATE <-as.data.frame(svyby(~POP_STATE, by=~state_postal, RECS, svytotal)) %>%
  rename(state = state_postal) %>%
  select(-se)
rownames(POP_STATE) <- NULL

HOTMA_STATE<-as.data.frame(svyby(~HOTMA, by=~state_postal, RECS, svytotal)) %>%
  rename(state = state_postal) %>%
  select(-se)
rownames(HOTMA_STATE) <- NULL

COLDMA_STATE<-as.data.frame(svyby(~COLDMA, by=~state_postal, RECS, svytotal)) %>%
  rename(state = state_postal) %>%
  select(-se)
rownames(COLDMA_STATE) <- NULL

merged <- POP_STATE %>%
  inner_join(HOTMA_STATE) %>%
  inner_join(COLDMA_STATE) %>%
  mutate(HOTMA_rate = HOTMA/POP_STATE) %>%
  mutate(COLDMA_rate = COLDMA/POP_STATE) %>%
  arrange(-HOTMA_rate)
```

HOTMA normalized by state population
```{r}
plot_HOTMA <- plot_usmap(data = merged, values = "HOTMA_rate") +
  scale_fill_continuous(low = "#f7f7f7", high = "#b2182b", na.value = "#f7f7f7", label = percent, name = "Heat-related illness prevalence (2020)", limits = c(0, 0.03), breaks = seq(0, 0.03, by = 0.01)) +
  theme(legend.position = "top")
```
2020 state level estimates
```{r}
plot_COLDMA <- plot_usmap(data = merged, values = "COLDMA_rate") +
  scale_fill_continuous(low = "#f7f7f7", high = "#2166ac", na.value = "#f7f7f7", label = percent, name = "Cold-related illness prevalence (2020)", breaks = seq(0, 0.03, by = 0.01), limits = c(0, 0.03)) +
  theme(legend.position = "top") 
```

```{r}
results_all <- results_2015 %>%
  full_join(results_2020) %>%
  mutate(year = factor(year)) %>%
  pivot_wider(names_from = variable, values_from = c(total, SE, CI)) %>%
  mutate(total_TEMPMA = total_HOTMA + total_COLDMA) %>%
  mutate(SE_TEMPMA = sqrt(SE_HOTMA^2 + SE_COLDMA^2)) %>%
  mutate(CI_TEMPMA = 1.96*SE_TEMPMA) %>%
  pivot_longer(cols = ends_with("MA"), names_to = c("metric", "variable"), names_sep = "_", values_to = "value") %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(label = case_when(variable == "TEMPMA" ~ "Any temperature",
                           variable == "HOTMA" ~ "Heat-related",
                           variable == "COLDMA" ~ "Cold-related"))
```

Population estimates plot
```{r}
plot_total <- ggplot(results_all, aes(x = reorder(label,total), y = total, group = year, fill = year)) +
  scale_fill_manual(values = my_pal2) +
  scale_y_continuous(label = comma) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.8) +
  geom_errorbar(aes(ymin = total-CI, ymax = total+CI), position = position_dodge(width = 0.8), width = 0.2, colour = "#000000", linewidth = 0.2) +
  annotate("text", x = "Heat-related", y = 2000000, label = "Total U.S. households\n2015: 118.2 million\n2020:123.5 million", size = 2) +
  theme_custom() +
  coord_flip() +
  ylab("Estimated number of U.S. households affected")
```

```{r}
plot_maps <- ggarrange(plot_HOTMA, plot_COLDMA, ncol = 2, nrow = 1)
ggarrange(plot_total, plot_maps, ncol = 1, nrow = 2, heights = c(1, 2))

ggsave("Figure1_PopulationEstimate.pdf", units = "in", width = 6.5, height = 5)
```

